### 2021-02-24 math/ia_2021/paper.tex
added
ia work

@@

% Suppresses headers, footers, and page numbers on title page
\begin{titlepage}
    \begin{center}
    \vspace*{4cm}
    A statistical approach to predicting a tsunami’s wave height \\
    \vspace{1cm}
    Can the height of a tsunami's wave be predicted based on the distance from and 
    the magnitude of a preceding earthquake? \\
    \vspace{4cm}
    Word count: TBD
    \vfill
    \vspace{0.1cm}
    \end{center}
    \end{titlepage}

@@

Naturally occurring disasters such as tsunamis are devastating for 
people and resources residing in risk areas. These tsunamis are caused 
by many different types of natural phenomena such as landslides, collapse of 
seamount, or more rarely, the impact of a meteorite. This paper however will focus 
on one of the more common cases of tsunami; one caused by the earthquakes. This 
is due to earthquakes causing around 72\% of earthquakes \cite{pacifictsunamimuseum}, and as the time between an 
earthquake event and its tsunami is relatively high compared to events such as landslides \cite{sue_nokes_walters}, 
leaving adequate time for prediction and correct preparation of the events properties, such as 
the wave height, as discussed in this paper.

@@

I have chosen to write about this subject as I have recently become interested by the 
statistics and its implications in real-life matters. This research has proved very 
interesting for me as it has abstracted the field of statistics for me. Furthermore, 
as I enjoy programming, the research I have done has helped me learn about the tools 
and technologies used by data analysts and other professionals of the matter.

\section{}

@@

\subsection{Processing the dataset for use}

The dataset, procured from the NOAA website \cite{noaa}, includes many tsunami 
runups, going back before the year 0. Due to this, the NOAA states uncertainties 
that can occur in its data. For example, reports that were recorded before the 
20th century tend to be based on written accounts, and not recorded seismic activity. 
Furthermore, the installation of a system called the Worldwide Standardized Seismograph 
Network in the 1960s allowed for more accurate data and for that reason any recorded 
activity before this date will be discarded from the dataset. The dataset itself can 
be accessed through two different methods; searching the database using the online 
tool or downloading it in a somewhat obscure Google Earth format. The online tool 
supports the download of search results in a much more human-readable \verb|.csv| format, 
so I search for all entries after the year 1900, and downloaded the result.

@@

The data must be then correctly split into a training and an test dataset. The training 
dataset will be used to generate a regression model while the test dataset will 
be used to calculate the general accuracy of the model, by predicting values and 
comparing those to the known value as found in the dataset. The dataset is split based on 
a decimal number using the \verb|df.sample| method found in the pandas library. A split 
of 70\% for training and 30\% for test has been chosen. A \verb|random_state| argument is 
set to an arbitrary value 42, as if it weren't, the dataset would be split in a different 
manner at each execution of the program.

@@

Regression analysis is a very large portion of statistics, and regression models can 
be found in numerous use cases of the field. Regression models consist of estimating the 
relationship between one (or more) independent variables and a dependent variable. A 
simple example of regression can be seen in the linear relationship between an independent 
variable $x_{i}$, and the dependent variable $y$. The following expression shows this relationship in what is 
called linear regression, as it models a straight line.

$y = a + \beta x + e $

Where the term $a$ represents the y-intercept of the line and $\beta$ represents the 
coefficient of $x_i$. The final term $e$ is what is known as the error or residual 
term, it is an estimate of the amount by which the predictions differ from the real 
value.

However, the relationship between the distance from an event and the wave height 
is not necessarily linear, especially if it is considered that the travel of 
sound in a fluid obeys the inverse square law. This already tells us a simple 
linear regression model cannot be used for this research. If we want a second degree 
polynomial regression formula, a new $x^2$ term can be added to the previous formula 
and we get:

@@

Even though the right side of this equation is quadratic in $x$, this is still 
linear regression as the right side is linear in $\beta$ (if we were to let 
$x = 1$, we would be left with $y = a + \beta_1 + \beta_2 + e$, which is linear). 
A second degree polynomial regression would best fit our model for both independent 
variables, as seen in the graphs below, where a theoretical regression line is drawn 
by hand:

@@

In addition, our dataset has two independent variables: the distance from a particular 
earthquake event and the magnitude of that earthquake.

@@

@@

The model's accuracy is then calculated, with the mentioned before "test" dataset. 
The \verb|accuracy_score| method from the \verb|sklearn.metrics| class is used for this 
task.

\subsection{Ease of access}
For an easy access to the predictions generated by the program, I implemented 
a simple function to fetch new earthquake events from the \url{https://earthquake.usgs.gov/earthquakes/feed/v1.0/geojson.php} 
live earthquake feed. These earthquake events are quantified into a magnitude \verb|mag| 
and a distance \verb|dist| variable, which are supplied to the statistical model. 
This information is then easily accessible from a human-readable report exported 
to a plaintext file. Following is an example report generated by the tool:

@@

When attempting to predict the maximum wave height of a tsunami such as the one 
caused by the 2011 earthquake off the Pacific coast of Tōhoku, approximately 70km 
away from shore, which produced a 38.9m wave \cite{yomiuri_2011}. The model in this 
paper approximates this same event as only a 8m wave, when supplied the 9.0 magnitude 
of the earthquake and the 70km distance. These types of errors can be attributed to the 
fact that there are many more independent variables, such as the shape of the seafloor for 
example, that can amplify or potentially change the height of a tsunami's wave. 
%todo add

@@

\appendix
\section{Appendix}
\label{app}
\subsection{Python program}
\label{app:scripts}
qsd
\lstinputlisting[language=Python]{model/multregr.py}

>>>>

\pagestyle{fancy}
\lhead{Väinö-verneri Kauppila} % controls the left corner of the header
\chead{} % controls the center of the header
\rhead{} % controls the right corner of the header
\lfoot{} % controls the left corner of the footer
\cfoot{} % controls the center of the footer
\rfoot{Page~\thepage} % controls the right corner of the footer
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}

@@

% Suppresses headers, footers, and page numbers on title page
\begin{titlepage}
    \begin{center}
        \vspace*{4cm}
        A statistical approach to predicting a tsunami’s wave height \\
        \vspace{1cm}
        Can the height of a tsunami's wave be predicted based on the distance from and
        the magnitude of a preceding earthquake? \\
        \vspace{1cm}
        \textit{Väinö-verneri Kauppila} \\
        February 2021 \\
        \vspace{4cm}
        Word count: TBD
        \vfill
        \vspace{0.1cm}
    \end{center}
\end{titlepage}

@@

Naturally occurring disasters, such as tsunamis, are devastating for people, human settlements, 
buildings and other resources residing in risk areas. Tsunamis are caused
by many different types of natural phenomena, such as landslides, collapse of
seamount, or more rarely, the impact of a meteorite. This paper, however, focuses
on one of the more common causes of tsunamis; earthquakes. Earthquakes cause around 72\% 
of tsunamis \cite{pacifictsunamimuseum}. The time between an earthquake and its tsunami is 
relatively long compared to events such as landslides \cite{sue_nokes_walters},
leaving adequate time to prepare for impact. Predicting wave height, as discussed in this 
paper, can further help to assess the seriousness of the tsunami and provide help to 
prepare for the event.

@@

I have chosen to write about this subject as I have recently become interested by the
statistics and its implications in real-life matters. This research has proved very
interesting for me as it has made the field of statistics more concrete, because it provides 
a real world use case. Furthermore, as I enjoy programming, the research I have done 
has helped me learn about the tools and techniques used by data analysts and other professionals.

@@

\subsection{Processing the dataset for use}
%TODO: #2 Ask for help 
The dataset, procured from the NOAA website \cite{noaa}, includes many tsunami
runups, going back before the year 0. Due to this, the NOAA states uncertainties
that can occur in its data. For example, reports that were recorded before the
20th century tend to be based on written accounts, and not recorded seismic activity.
Only since the 1960s, the installation of the Worldwide Standardized Seismograph
Network system allowed for more accurate data in this paper. For that reason, any recorded
activity before this date was discarded from the dataset for this paper. The dataset itself can
be accessed through two different methods; searching the database using the online
tool or downloading it in a somewhat obscure compressed Google Earth \verb|.kmz| format. 
The online tool supports the download of search results in a much more human-readable 
\verb|.csv| format. For this paper, all entries after the year 1900 were searched and downloaded.

@@

The data was split into a training and a test dataset. The training
dataset was used to generate a regression model while the test dataset was used 
to calculate the general accuracy of the model, by predicting values and
comparing those to the known value as found in the dataset. The dataset was split based on
a fraction using the \verb|df.sample| method found in the pandas library. A split
of 70\% for training and 30\% for test was chosen. A \verb|random_state| argument is
set to an arbitrary value 42, as if it weren't, the dataset would be split in a different
manner at each execution of the program.

@@

Regression analysis is a commonly used tool in statistics, and regression models can
be found in numerous use cases. Regression models consist of estimating the
relationship between one (or more) independent variables and a dependent variable. A
simple example of regression can be seen in the linear relationship between an independent
variable $x_{i}$, and the dependent variable $y$. The following expression shows this relationship in what is
called linear regression, as it models a straight line.

$y = a + \beta x + e $,

where the term $a$ represents the y-intercept of the line and $\beta$ represents the
coefficient of $x_i$. The final term $e$ is what is known as the error or residual
term. It is an estimate of the amount by which the predictions differ from the real
value.

However, the relationship between the distance between an earthquake and its tsunami 
and the wave height is not necessarily linear, especially if it is considered that the travel of
sound (which is a wave, or the displacement of a fluid that can travel) in a fluid obeys the 
inverse square law. Therefore, a simple linear regression model cannot be 
used for this research. A second degree polynomial regression formula would be more 
appropriate, and a new $x^2$ term can be added to the previous formula. We get:

@@

Even though the right side of this equation is quadratic in $x$, this is still
linear regression as the right side is linear in $\beta$ (if we were to let
$x = 1$, we would be left with $y = a + \beta_1 + \beta_2 + e$, which is linear).
A second degree polynomial regression would best fit our model for both independent
variables, as seen in the graphs below, where a theoretical curve of regression is drawn
by hand:

@@

In addition, our dataset has two independent variables: the distance from a particular
earthquake event and the magnitude of that earthquake. Models such as the one explored 
in this paper, where multiple independent variables are present, are called multivariate 
and can be modeled by adding the terms of a new independent variable, as such 
for a linear model with two independent variables:

$a_1 + \beta_1 x_1 + a_2 + \beta_2 x_2 + e$,

where the terms $a_n$ are the y-intercepts of the $n$th independent variable,

\begin{figure}[h]
    \centering
    \subfloat[\centering Linear model]{{\includegraphics[width=5cm]{linear.png} }}
    \qquad
    \subfloat[\centering Polynomial model]{{\includegraphics[width=5cm]{poly.png} }}
    \caption{2 multivariate (hence the 3 dimensions) regression models}
    \label{fig:example}
\end{figure}

@@

The model's accuracy is then calculated, with the test dataset. At first, it was considered 
to use the \verb|accuracy_score| method from the \verb|sklearn.metrics| class was used for 
this task. This function calculates what is called the $R^2$ value, which is a measure used to 
evaluate the fit a model on the dataset.

@@

@@

When attempting to predict the maximum wave height of a tsunami such as the one
caused by the 2011 earthquake off the Pacific coast of Tōhoku, approximately 70km
away from shore, which produced a 38.9m wave \cite{yomiuri_2011}. The model in this
paper approximates this same event as only a 8m wave, when supplied the 9.0 magnitude
of the earthquake and the 70km distance. These types of errors can be attributed to the
fact that there are many more independent variables, such as the shape of the seafloor for
example, that can amplify or potentially change the height of a tsunami's wave.

@@

\appendix
\section{Appendix}
\label{app}
\subsection{Python program}
\label{app:scripts}

Listed here is the Python program through which the dataset was cleaned and
predictions based on the statistical model were made. Run using 64-bit Python
version 3.9.1.

\lstinputlisting[language=Python]{model/multregr2.py}
